---
title: "R Notebook"
output: html_notebook
---

Disability Employment Awareness
B18120 Employment Status by Disability Status and Type (Civilian Noninstitutionalized Population 18 to 64 Years)
B18121 Work Experience by Disability Status and Type
B18130 Age by Disability Status by Poverty Status

[Census Stats](https://www.census.gov/newsroom/stories/disability-employment-awareness-month.html)

- stats by age, sex, and disability type -- can apply for just our region


Previous blog:

- people with disability (without) in the region
- low incomes compared to those in the regions general population
- unemployment rate for people with disabilities compared to population overall
- people of color with disabilities and low income (+ regional average)
- regional ORCA card, ridership of people with disabilities

Check for disability topics in HHTSE

Sources:
[Coordinated Mobility Plan](https://www.psrc.org/media/6395)
[Heritage Month Blog, 2022](https://www.psrc.org/about-us/media-hub/helping-people-all-abilities-get-work)
Census pulse data (processed data Seattle MSA)

```{r}
library(psrccensus)
library(psrcplot)
library(tidyverse)
library(tidycensus)
```

```{r, B18120 Employment Status}
x <- get_acs_recs('county', table.names = 'B18120', years = '2022', acs.type = 'acs1')
# x_fmt <- format_estimates(x)

x02 <- x %>% 
  mutate(variable_sort = str_extract(variable, "\\d{2}$")) %>% 
  mutate(label02 = label) %>% 
  separate(label02,
           sep = ":!!",
           into = c('total', 'in_lf', 'is_emp', 'has_dis', 'dis_type')) %>%
  mutate(dis_type = ifelse(variable_sort %in% 23:28, has_dis, dis_type)) %>%
  mutate(has_dis = ifelse(variable_sort %in% 22:29, is_emp, has_dis)) %>%
  mutate(is_emp = ifelse(variable_sort %in% 22:29, NA, is_emp)) %>% 
  mutate(in_lf = str_replace_all(in_lf, ":", ""),
         is_emp = str_replace_all(is_emp, ":", ""),
         has_dis = str_replace_all(has_dis, ":", ""))
 

  
# region ----

r <- x02 %>% 
  filter(name == 'Region')

r_lf <- r %>% 
  filter(!is.na(in_lf)) %>% 
  group_by(name, in_lf) %>% 
  summarise(estimate = sum(estimate), 
            moe = moe_sum(moe = moe, estimate = estimate)) %>% 
  mutate(se = moe/1.645) %>% 
  mutate(cv = (se/estimate)*100)
  


```

```{r}

create_shares <- function(table) {
  df <- format_estimates(table, type = 'share')
  df <- df %>% 
    rename_with(~str_replace_all(.x, 'share_moe', 'moe'), ends_with('share_moe')) 
  
  df <- df %>% 
    pivot_longer(cols = King_share:Region_moe,
                 names_to = c('county', "type"),
                 names_sep = "_") %>% 
    pivot_wider(names_from = type,
                values_from = value) %>% 
    arrange(county, variable)
  return(df)
}

y <- create_shares(x)
```

